Equivalence Hypergraphs: E-Graphs for Monoidal Theories

### 2024-01-18 E-Hypergraphs-LICS24.tex
finish S5

    \tikzfig{combinatorial_semantics/DPOI_pushout_complement}
}
\]
is a \textit{boundary} complement if
\begin{enumerate} 
    \item the image of $m$ is a convex down-closed e-hypergraph
    \item $[c_1,c_2]$ is mono
    \item there exist $d_1 : n \to \mathcal{L}^\bot$ and $d_2 : k \to \mathcal{L}^\bot$ making the above triangle commute
    \item all $v,w$ in $\mathcal{G}$ with the pre-image in $i + j$ share the same set of parents
    \item either for all $v,w$ in $\mathcal{G}$ with the pre-image in $i+j$, $v \consistency w$, or for all $v,w$ in $\mathcal{G}$ with the pre-image in $i+j$ $\consistency$ is undefined
    \item if the image of $\mathcal{L}$'s external interfaces under $m$ consists exclusively of top-level vertices of $\mathcal{G}$ then there exists a \textit{well-typed} extended mda-cospan
    \[
    n + j \xrightarrow{f_1 + id_{j}} n' \setminus (i' \setminus i) + j \xrightarrow{[g_1,c_2]} \mathcal{L}^{\bot} \xleftarrow{[g_2,c_1]} k' \setminus (j' \setminus j) + i \xleftarrow{f_2 + id_{i}} k + i
    \]
    \item if the image of $\mathcal{L}$'s external interfaces under $m$ consists exclusively of not top-level vertices of $\mathcal{G}$ then there exists a \textit{not necessarily well-typed} extended mda-cospan
    \[
    n \xrightarrow{f_1} n' \setminus (i' \setminus i) + j \xrightarrow{[g_1,c_2]} \mathcal{L}^{\bot} \xleftarrow{[g_2,c_1]} k' \setminus (j' \setminus j) + i \xleftarrow{f_2} k
    \]
\end{enumerate}

We first note that because the image of $m$ is a convex down-closed e-hypergraph, $i' \setminus i$, which is a strict internal interface of $\mathcal{L}$, is necessarily mapped to vertices in the strict internal interface of $\mathcal{G}$ which are also in the image of $n'$ and therefore we can construct a morphism $f' : i' \setminus i \to n'$ by first following $m$ and then composing with the partial inverse of $g_{int}$ which exists because it is mono. 
Then we define $n' \setminus (i' \setminus i)$ as the pushout complement in the diagram below

\[
    \tikzfig{combinatorial_semantics/pushout_difference}
\]
and $g_1$ is obtained by composition $g';g_{int}$ and then composing with the inverse of $l$.
Similarly for $g_2$.
$f_1$ is given by $g_{ext}$ with its codomain restricted to $n' \setminus (i' \setminus i)$ --- noting the image of $g_{ext}$ lies within $n' \setminus (i' \setminus i)$.
Similarly for $f_2$.
% CHRIS: TRY TO DEFINE G1, G2 AS SIMPLY AS POSSIBLE, AND POSSIBLY ADD AN INFORMAL EXPLANATION BELOW THE DEFINITION. 
% Because $\mathcal{L}^{\bot}$ is essentially $\mathcal{G}$ with $\mathcal{L}$ removed, apart from the part that has a pre-image in $i + j$, there is an obvious morphism from $l \to \mathcal{L}^{\bot}$ which is $g_1 = g_{int} |_{D}$ where
% \[
%     D = \{ v \in n' \text{ such that } \not \exists u \in (i' \setminus i) ~ . ~ g_{int}(v) = f'_{int};f(u) \}
% \]
% Similarly for $r \to \mathcal{L}^{\bot}$, which we denote as $g_2$.
% Then there exists a morphism $n \to l$, because there is a morphism $n \to n'$ and the external interfaces of $n'$ and $l$ are the same.
% That is, it is a morphism $f: n \to l$ such that $d_1 = f;g_1$.
% Similarly for $k \to r$.
\end{definition}

@@

\begin{definition}[Convex EDPOI rewriting]
\label{def:dpoi-e}
Given an extended span of morphisms 
\[
    \mathcal{G} \xleftarrow{} n' + k' \xleftarrow{} n + k \xrightarrow{} n'' + k'' \xrightarrow{} \mathcal{H}
\]
in $\catname{EHyp(\Sigma)}$, we say $\mathcal{G}$ \textit{rewrites (convexly) to} $\mathcal{H}$ (\textit{with external interface} $n + k$ and \textit{taking internal interface} $n'+k'$ \textit{to} $n'' + k''$) --- denoted by $\mathcal{G} \Rrightarrow \mathcal H$  --- \textit{via a rewrite rule} 
\[
    \mathcal{L} \xleftarrow{} i' + j' \xleftarrow{} i + j \xrightarrow{} i'' + j'' \xrightarrow{} \mathcal{R}
\] 
if there exists an object $\mathcal{C}$ and morphisms which complete the following commutative diagram 
\[
 \scalebox{0.75}{
    \tikzfig{combinatorial_semantics/DPOI_square}
 }
\]
such that the two marked squares are pushouts and the following conditions hold:
    \begin{enumerate}
        \label{dpoi-e:assumptions}
        \item $m : \mathcal{L} \to \mathcal{G}$ is a convex match;
        \item $m(L)$ is a down-closed e-hypergraph;
        \item $i + j \to \mathcal{C} \to \mathcal{G}$ is a boundary complement;
	\item the internal interfaces of $\mathcal H$ are such that:
	\[
		n'' = n' \setminus (i' \setminus i) + (i'' \setminus i) \qquad k'' = k' \setminus (j' \setminus j) +  (j'' \setminus j)\\
	\]
		\begin{align*}
		&f_1 : i'' \setminus i \to \mathcal{R} \to \mathcal{H} & \\ 
		&f_2 : n' \setminus (i' \setminus i) \to \mathcal{C} \to \mathcal{H} & \\
		\end{align*}
    \end{enumerate}
% The internal interfaces of the resulting e-hypergraph $\mathcal{H}$ from the cospan
%  \[
%     \mathcal{H} \xleftarrow{} n'' + k'' \xleftarrow{} n + k
% \]

$[f_1,f_2]: n'' \to \mathcal{H} $
Since both $f_1$ and $f_2$ are monos with disjoint image (?), their copairing is also a mono.  Note $f_1$ is the obvious restriction,  $f_2$ is defined as $g_2$ in boundary complement. 
CHRIS: FIX (4) ABOVE:

    Given  a set $\mathfrak{R}$ of EDPOI rewrite rules and e-hypergraphs with extended interfaces $\mathcal G$ and $\mathcal H$,  we write $\mathcal G \Rrightarrow_\mathfrak{R} \mathcal H$ if there exists a EDPOI rewrite in $\mathfrak R$ such that via it $\mathcal G$ rewrites to $\mathcal H$,  and we write $\mathcal G \Rrightarrow^*_\mathfrak{R} \mathcal H$ if there exists a sequence of such rewrites. 
\end{definition}
The boundary complement conditions, in particular, prohibits finding a match for $\mathcal L$ in $\mathcal G$,  below.  
\[
	\mathcal{L} = \scalebox{0.8}{\tikzfig{combinatorial_semantics/f_times_g}} \qquad \mathcal{G} = \scalebox{0.6}{\tikzfig{combinatorial_semantics/f_plus_g_inline}}
\]
Together, the conditions ensure the result of a rewrite is a well-typed MDA e-hypergraph.

>>>>

    \tikzfig{combinatorial_semantics/DPOI_pushout_complement}
}
\]
is a \textit{boundary complement} if
\begin{enumerate} 
    \item $m(\mathcal L)$ is a convex down-closed e-hypergraph;
    \item $[c_1,c_2]$ is mono;
    \item for all $v,w$ in $\mathcal{G}$ in the image of $i + j$,  $v$ and $w$ share the same set of parents, and either $v,w$ are top-level or else $v \consistency w$;
    \item there exist $d_1 : n \to \mathcal{L}^\bot$ and $d_2 : k \to \mathcal{L}^\bot$ making the above triangle commute; and
    \item if the image of $\mathcal{L}$'s external interfaces under $m$ consists exclusively of top-level vertices of $\mathcal{G}$ then there exists a \textit{well-typed} extended mda-cospan
    \[
    n + j \xrightarrow{f_1 + id_{j}} n' \setminus (i' \setminus i) + j \xrightarrow{[g_1,c_2]} \mathcal{L}^{\bot} \xleftarrow{[g_2,c_1]} k' \setminus (j' \setminus j) + i \xleftarrow{f_2 + id_{i}} k + i
    \]
    \item if the image of $\mathcal{L}$'s external interfaces under $m$ consists exclusively of not top-level vertices of $\mathcal{G}$ then there exists a \textit{not necessarily well-typed} extended mda-cospan
    \[
    n \xrightarrow{f_1} n' \setminus (i' \setminus i) + j \xrightarrow{[g_1,c_2]} \mathcal{L}^{\bot} \xleftarrow{[g_2,c_1]} k' \setminus (j' \setminus j) + i \xleftarrow{f_2} k
    \]
\end{enumerate}
where $f_1$ is given by $g_{ext}$ with its codomain restricted to $n' \setminus (i' \setminus i)$ --- noting the image of $g_{ext}$ indeed lies within $n' \setminus (i' \setminus i)$ --- and 
similarly for $f_2$. The map $g_1$ is derived from $g_{int}$ as follows. 
 %because the image of $m$ is a convex down-closed e-hypergraph, 
Note that the internal interface $i'$ of $\mathcal L$ is mapped to the internal interface $n'$ of $\mathcal G$,  since $\mathcal L$ is a down-closed subgraph of $\mathcal G$.  This induces an identification of $i'$ in  $n'$ which restricts to an identification of $i'\setminus i$ in $n'$. 
This gives rise to a restriction of $g_{int}$ of type $n' \setminus (i' \setminus i) \to \mathcal G$.  Since $\mathcal L^\bot$ can be identified within $\mathcal G$ --- and in particular has the internal interface of $\mathcal G$ minus $i' \setminus i$ --- this observation induces the map $g_1:  n' \setminus (i' \setminus i) \to \mathcal{L}^\bot$ from $g_{int}$.  A similar argument defines $g_2$. 
%$i' \setminus i$, which is a strict internal interface of $\mathcal{L}$, is necessarily mapped to vertices in the strict internal interface of $\mathcal{G}$ which are also in the image of $n'$ and therefore we can construct a 
%morphism $f' : i' \setminus i \to n'$ by first following $m$ and then composing with the partial inverse of $g_{int}$ which exists because it is mono. 
%Then we define $n' \setminus (i' \setminus i)$ as the pushout complement in the diagram below
%
%\[
%    \tikzfig{combinatorial_semantics/pushout_difference}
%\]
%and $g_1$ is obtained by composition $g';g_{int}$ and then composing with the inverse of $l$.
%Similarly for $g_2$.
% CHRIS: TRY TO DEFINE G1, G2 AS SIMPLY AS POSSIBLE, AND POSSIBLY ADD AN INFORMAL EXPLANATION BELOW THE DEFINITION. 
% Because $\mathcal{L}^{\bot}$ is essentially $\mathcal{G}$ with $\mathcal{L}$ removed, apart from the part that has a pre-image in $i + j$, there is an obvious morphism from $l \to \mathcal{L}^{\bot}$ which is $g_1 = g_{int} |_{D}$ where
% \[
%     D = \{ v \in n' \text{ such that } \not \exists u \in (i' \setminus i) ~ . ~ g_{int}(v) = f'_{int};f(u) \}
% \]
% Similarly for $r \to \mathcal{L}^{\bot}$, which we denote as $g_2$.
% Then there exists a morphism $n \to l$, because there is a morphism $n \to n'$ and the external interfaces of $n'$ and $l$ are the same.
% That is, it is a morphism $f: n \to l$ such that $d_1 = f;g_1$.
% Similarly for $k \to r$.
\end{definition}

@@

\begin{definition}[Convex EDPOI rewriting]
\label{def:dpoi-e}
Given an extended span of morphisms 
\[
    \mathcal{G} \xleftarrow{} n' + k' \xleftarrow{} n + k \xrightarrow{} n'' + k'' \xrightarrow{} \mathcal{H}
\]
in $\catname{EHyp(\Sigma)}$, we say $\mathcal{G}$ \textit{rewrites (convexly) to} $\mathcal{H}$ (\textit{with external interface} $n + k$ and \textit{taking internal interface} $n'+k'$ \textit{to} $n'' + k''$) --- denoted by $\mathcal{G} \Rrightarrow \mathcal H$  --- \textit{via a rewrite rule} 
\[
    \mathcal{L} \xleftarrow{} i' + j' \xleftarrow{} i + j \xrightarrow{} i'' + j'' \xrightarrow{} \mathcal{R}
\] 
if there exists an object $\mathcal{C}$ and morphisms which complete the following commutative diagram 
\[
 \scalebox{0.75}{
    \tikzfig{combinatorial_semantics/DPOI_square}
 }
\]
such that the two marked squares are pushouts and the following conditions hold:
    \begin{enumerate}
        \label{dpoi-e:assumptions}
       % \item $m : \mathcal{L} \to \mathcal{G}$ is a convex match;
        %\item $m(L)$ is a down-closed e-hypergraph;
        \item $i + j \to \mathcal{C} \to \mathcal{G}$ is a boundary complement;
	\item the internal interfaces of $\mathcal H$ are such that:
	\[
		n'' = n' \setminus (i' \setminus i) + (i'' \setminus i) \qquad k'' = k' \setminus (j' \setminus j) +  (j'' \setminus j)\\
	\]
	\item the map $f = [f_1,f_2]: n'' \to \mathcal H$ in the diagram above is defined as in Definition \ref{def:boundary_new} of boundary complement,  and $f_2$ as the restiction of the composite $i'' \to \mathcal R \to \mathcal H$ and,  and similarly for the map $g: k'' \to \mathcal H$.  
    \end{enumerate}
% The internal interfaces of the resulting e-hypergraph $\mathcal{H}$ from the cospan
%  \[
%     \mathcal{H} \xleftarrow{} n'' + k'' \xleftarrow{} n + k
% \]
    Given  a set $\mathfrak{R}$ of EDPOI rewrite rules and e-hypergraphs with extended interfaces $\mathcal G$ and $\mathcal H$,  we write $\mathcal G \Rrightarrow_\mathfrak{R} \mathcal H$ if there exists a EDPOI rewrite in $\mathfrak R$ such that via it $\mathcal G$ rewrites to $\mathcal H$,  and we write $\mathcal G \Rrightarrow^*_\mathfrak{R} \mathcal H$ if there exists a sequence of such rewrites. 
\end{definition}
The boundary complement conditions, in particular, prohibits finding a match for $\mathcal L$ in $\mathcal G$,  below.  
\[
	\mathcal{L} = \scalebox{0.8}{\tikzfig{combinatorial_semantics/f_times_g}} \qquad \mathcal{G} = \scalebox{0.6}{\tikzfig{combinatorial_semantics/f_plus_g_inline}}
\]
Together, the conditions ensure the result of a rewrite is a well-typed MDA e-hypergraph.
